---
title: "Lab 5"
author: "Nicholas Heger"
date: "11/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggrepel)
library(maps)
library(ggthemes)
states <- map_data("state")
counties <- map_data("county")
acc <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv", stringsAsFactors = FALSE)
per <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv", stringsAsFactors = FALSE)
unitedStates <- readxl::read_xlsx('./FRPP_GLC_-_United_StatesSep292020.xlsx')
dayCodes <- readxl::read_xlsx("./dayCodes.xlsx")
makeCodes <- read.csv("./makeCodes.csv")
stateCodes <- read.csv("./stateCodes.csv")
```

## Questions
1. Are there some days of the week where more accidents happen than the others (see FARS manual, use variable DAY_WEEK)?  
```{r}
acc1 <- acc %>%
  group_by(DAY_WEEK)
count(acc1)
ggplot(acc, aes(x=DAY_WEEK)) + geom_bar() + ggtitle('Number of Accidents per Day of Week')
```
  
Yes, accidents happen most often on Saturdays. I confirmed that 7 corresponds to Saturday by checking the user's manual for the data.  

2. Create a data frame containing the persons who are fatally hurt in the accidents (look up variable INJ_SEV)  
```{r}
fatal <- per %>% 
  filter(INJ_SEV == 4)
head(fatal)
```
  
3. Create a data frame containing the most dangerous vehicle make in each state. The number of persons fatally hit in the vehicle make is used to assess the (non-)safety of a make. Make sure to handle the missing values appropriately. (look up variable MAKE)  
```{r}
FatalMakeState <- fatal %>% 
  group_by(STATE, MAKE)
count1 <- count(FatalMakeState)
badMakes <- count1 %>% 
  arrange(STATE, desc(n))
badMakes <- badMakes[complete.cases(badMakes),]
bad <- badMakes %>% 
  group_by(STATE) %>% 
  slice(1)
bad
```
  
4. Create a map, and label each state with the most dangerous vehicle. Discuss the definition of the most dangerous vehicle, and what you find from the map. (Hint: Read the description for the STATE and COUNTY columns in the FARS manual. The state & county codes are Geographic Locator Codes (GLCs) from the General Services Administration’s (GSA) publication. Use readxl::read_xlsx to read in the GLCs.)
```{r}
stateCodes$State <- tolower(stateCodes$State)
badMakesWithCodes <- bad %>%
  left_join(makeCodes, by=c('MAKE'='Codes')) %>%
  left_join(stateCodes, by=c('STATE'='Codes'))
badMakesWithCodesMap <- badMakesWithCodes %>%
  left_join(states, by=c('State'='region'))
states %>%
  group_by(region) %>%
  summarise(long=mean(long), lat=mean(lat), .groups='keep') %>%
  left_join(badMakesWithCodes, by=c('region'='State')) -> stateLabels

ggplot(badMakesWithCodesMap, aes(x=long, y=lat)) +
  geom_path(aes(group=group)) +
  geom_text_repel(data=stateLabels, aes(label=Make), size=3) +
  ggtitle('Most Dangerous Make of Car by State') +
  theme_map()
```
  
The most dangerous makes being shown in this graph are defined by the make of car that the most fatal injuries have occurred in. Because this is not a percentage, the makes shown here are more likely to be more popular makes of cars since more people own them. We can find from this map that Chevrolet seems to be a dangerous (and/or popular) make of car in most of the united states, especially the Midwest.  

5. Join the accident and person table (work out which variable(s) to use)  
```{r}
acc_per <- inner_join(acc, per, by=c("STATE", "ST_CASE", "VE_FORMS", "COUNTY", "DAY", "MONTH", "HOUR", "MINUTE", "RUR_URB", "FUNC_SYS", "HARM_EV", "MAN_COLL", "SCH_BUS"))
str(acc_per)
```
  
6. Tally the number of accidents by day of the week (DAY_WEEK), hour of the day (HOUR) and gender (SEX). Visualize the results.
```{r}
accidentsByDayHourGender <- acc_per %>% 
  group_by(DAY_WEEK, HOUR, SEX) %>% 
  summarize(Number_of_Accidents = sum(DAY_WEEK, HOUR, SEX)) %>% 
  filter((SEX == 1 | SEX == 2) & HOUR != 99) %>% 
  mutate(SEX = ifelse(SEX == 1, 'Male', 'Female'))
accidentsByDayHourGender <- left_join(accidentsByDayHourGender, dayCodes, by=c('DAY_WEEK' = 'Code'))
accidentsByDayHourGender
accidentsByDayHourGender %>% ggplot(aes(x=HOUR, y = Number_of_Accidents, fill=SEX)) + geom_bar(stat="identity") + facet_wrap(~Day_Week) + ylab('Accidents') + xlab('Hour of Day') + ggtitle('Accidents by day of week, hour of day, and gender')
```
  
According to the above graph, more males are involved in accidents than females, and Friday and Saturday seem to be the days with the most accidents, mostly in the evenings and nights.  

7. Now plot a choropleth map of the number of deaths on a county level.
```{r}
countiesTemp <- counties
accidentsByCounty <- acc_per %>% 
  group_by(STATE, COUNTY, INJ_SEV) %>% 
  filter(INJ_SEV == 4) %>% 
  summarize(total = sum(INJ_SEV))
countyCodes <- unitedStates %>% 
  select('County Name', 'County Code', 'State Name', 'State Code') %>% 
  mutate('County Name' = tolower(unitedStates$`County Name`)) %>% 
  mutate('State Name' = tolower(unitedStates$`State Name`))
countiesTemp <- right_join(countiesTemp, countyCodes, by=c("region" = "State Name", "subregion" = "County Name"))
countiesTemp <- countiesTemp %>% 
  mutate("County Code" = as.numeric(countiesTemp$`County Code`)) %>% 
  mutate("State Code" = as.numeric(countiesTemp$`State Code`)) %>% 
  left_join(accidentsByCounty, by=c("County Code" = "COUNTY", "State Code" = "STATE")) %>% 
  mutate(total.x = ifelse(is.na(total), 0, total))
ggplot(countiesTemp, aes(x=long, y=lat, fill=total.x)) + geom_polygon(aes(group = group, fill=total.x)) + coord_map() + ggtitle('Car Accident Deaths by County')
```
  
The above graph shows that the counties with the most car accident deaths are in Southern California, with a few other higher car accident death counties in Texas and Florida.  

8. Is summer or winter more dangerous? Does this depend on states? Explore and explain.  

Summer
```{r}
summerFatalities <- fatal %>%
  filter(MONTH >= 6 & MONTH <=9)
summerFatalitiesByState <- summerFatalities %>%
  select(ST_CASE, STATE, PER_NO) %>%
  group_by(STATE) %>%
  summarise(Fatalities=sum(PER_NO))
summerFatalitiesByState
mean(summerFatalitiesByState$Fatalities)
```
  
Winter  
```{r}
winterFatalities <- fatal %>%
  filter(MONTH <= 3 | MONTH == 12)
winterFatalitiesByState <- winterFatalities %>%
  select(ST_CASE, STATE, PER_NO) %>%
  group_by(STATE) %>%
  summarise(Fatalities=sum(PER_NO))
winterFatalitiesByState
mean(winterFatalitiesByState$Fatalities)
```
  
The above data shows that summer months tend to have more fatal car accidents that winter months do. It also appears that more southern states tend to have more fatal accidents in the winter, while northern states do in the summer. I was initially surprised that more fatal car accidents occurred in the summer, but after thinking about why I believe this is because more people are likely to travel in the warm summer months than in the winter.
